---
title: "convert to anndata"
---

```{r}
library(Seurat)
library(SeuratDisk)
library(SeuratObject)
#library(MuDataSeurat)
library(reticulate)

#use_condaenv("cell2location", required = T)

# check
#py_config()
```

```{r}
integrated = qs::qread("rdata/02_04_integrated.qs")

DefaultAssay(integrated) = "RNA"

# diet, cannot remove scaled.data through layers argument! bug! 
diet = DietSeurat(
  integrated,
  features = rownames(integrated),
  assays = "RNA",
  dimreducs = c("rpca.rnaPCA.wnnUMAP"),
  graphs = c("rpca.rnaPCA.wsnn","rpca.rnaPCA.wknn")
)

# create RNA v3 assay, MANDATORY for writing to h5ad
diet[["RNA"]] = as(object = diet[["RNA"]], Class = "Assay")

# drop scaled data here!!! NEED TO DROP SCALED DATA or else counts will not be preserved
diet[["RNA"]]$scale.data <- NULL


# convert table into characters otherwise only preserves levels!
meta = diet@meta.data

meta = data.frame(lapply(meta, as.character), stringsAsFactors=FALSE)

diet@meta.data = meta
```

-   

    -   `X` will be filled with `scale.data` if `scale.data` is present; otherwise, it will be filled with `data`

    -   `var` will be filled with `meta.features` **only** for the features present in `X`; for example, if `X` is filled with `scale.data`, then `var` will contain only features that have been scaled

    -   `raw.X` will be filled with `data` if `X` is filled with `scale.data`; otherwise, it will be filled with `counts`.
        If `counts` is not present, then `raw` will not be filled

    -   `raw.var` will be filled with `meta.features` with the features present in `raw.X`; if `raw.X` is not filled, then `raw.var` will not be filled

-   <https://mojaveazure.github.io/seurat-disk/reference/Convert.html>

-   TO DO: can you save multiple slots or is it mu data type only?

```{r}
SaveH5Seurat(
  diet,
  filename = "/scratch/kou/composite/rdata/forVal_all_RNA.h5Seurat",
  overwrite = T
)

Convert(
  source = "/scratch/kou/composite/rdata/forVal_all_RNA.h5Seurat",
  dest = "h5ad",
  overwrite = T, 
  assay = "RNA"
)
```
