---
title: "QC of composite lymphoma 10X multiome data: GEX"
---

**Packages**

```{r}
#| label: setup

library(Seurat)
library(tidyverse, quietly = T)
library(reticulate)
library(scDblFinder)
library(SoupX)
library(DropletUtils)
library(furrr)
library(future.apply)
library(scCustomize)
library(ggridges)

theme_set(theme_bw())

# always set seed!!!
set.seed(105)

# future settings
plan(list(
  tweak(multisession, workers = availableCores() %/% 4),
  sequential
  )
)
```

## GEX data import

-   Individual modalities are preprocessed separately prior to integration.
-   Import both raw and filtered features to perform SoupX.
-   TO DO: use cellbender to skip SoupX and proceed to

```{r}
#| label: data_import

# list of samples
ln_id = c("LN0025","LN0027","LN0177","LN0193","LN0438")

# import droplets (raw) ----
droplets = paste0(
  "~/composite/cellranger/",
  ln_id,
  "/outs/raw_feature_bc_matrix/"
)

list_drop = future_lapply(droplets,Read10X)
names(list_drop) = ln_id

# keep only GEX
list_drop = future_map(list_drop, ~ .x$`Gene Expression`)

# import filtered count ----
nondrops = paste0(
  "~/composite/cellranger/",
  ln_id,
  "/outs/filtered_feature_bc_matrix/"
)

list_nondrop = future_lapply(nondrops,Read10X)
names(list_nondrop) = ln_id

# keep only GEX
list_nondrop = future_map(list_nondrop, ~ .x$`Gene Expression`)
```

## Droplet processing

-   Want to determine which barcodes are cells vs empty droplets
-   Want to remove ambient mRNA from empty droplets later

### Knee plot

```{r}
#| label: barcode_ranking
bcranks = future_lapply(list_drop, barcodeRanks)

# write function to plot
plot_knee = function(bc_out){

  # get knee/inflection data
  knee = metadata(bc_out)$knee
  infl = metadata(bc_out)$inflection
  
  p = as_tibble(bc_out) %>% 
    ggplot(., aes(rank,total))+
    geom_jitter(size = 0.5) +
    geom_hline(yintercept = c(knee,infl), color = c("blue","forestgreen")) +
    scale_x_continuous(trans = "log10") +
    scale_y_continuous(trans = "log10") +
    labs(x = "Barcode Rank", y = "UMI Counts")
  
  return(p)
}

knees = future_map(
  seq_along(bcranks),
  ~ plot_knee(bcranks[[.x]])
)

knplt = cowplot::plot_grid(plotlist = knees, labels = ln_id)
ggsave("figures/01_01_knee.png", knplt, width = 12, height = 8)
```

## SoupX removal of ambient RNA

### Creating `SoupChannel` objects

-   input required: table of droplets and table of counts
-   also need to add clustering and dimensionality reduction embeddings to metadata
-   can use `cellranger` preliminary analysis outputs

```{r}
#| label: createSoupChannelObject
list_soup = future_map2(
  list_drop, 
  list_nondrop, 
  ~ SoupChannel(.x,.y),
  .options = furrr_options(scheduling = Inf)
)

# get clustering data, use cell ranger output
clusterPath = paste0(
  "~/composite/cellranger/",
  ln_id,
  "/outs/analysis/clustering/gex/graphclust"
)

clusterFiles = list.files(
  path=clusterPath, 
  pattern="clusters.csv", 
  full.names=T
)

list_clusters = lapply(clusterFiles, read_csv)
names(list_clusters) = ln_id

# add clusters to soup channel
# checked all(rownames(list_soup$LN0025$metaData) == list_clusters$LN0025$Barcode)
list_soup = future_map2(
  list_soup, 
  list_clusters, 
  ~ setClusters(.x,.y$Cluster),
  .options = furrr_options(scheduling = Inf)
)

# get umap embeddings
umapPath = paste0(
  "~/composite/cellranger/",
  ln_id,
  "/outs/analysis/dimensionality_reduction/gex"
)

umapFiles = list.files(path = umapPath, pattern="umap_projection.csv", full.names=T)

list_umap = lapply(umapFiles, read_csv)
names(list_umap) = ln_id

# add umap embeddings to soup channel
list_soup = map2(
  list_soup, 
  list_umap, 
  ~ setDR(.x,.y[c("UMAP-1","UMAP-2")],"UMAP")
)
```

### Visualization of contamination

Before soup contamination removal: check expression of cell-specific markers (count \> 1)

```{r ms4a1_beforeCorrection}
# create function
featureUMAP = function(soupchannelobj,gene){
  ggplot(soupchannelobj$metaData, aes(UMAP_1,UMAP_2))+
    geom_jitter(
      aes(color = soupchannelobj$toc[gene,]>1), 
      size=0.1,
      alpha = 0.5
    )+
    theme(aspect.ratio = 3/3,
          panel.grid = element_blank())+
    labs(color = toString(gene))
}

cowplot::plot_grid(plotlist = lapply(list_soup,featureUMAP, gene = "TRAC"))
```

Plot to see if expression is more than expected in soup.
In the "CD247" (CD3 zeta) example, can see that the B cell clusters do not overcome threshold above soup!

```{r markermap}
cowplot::plot_grid(plotlist = lapply(list_soup, plotMarkerMap, geneSet = "CD247", pointSize = 1))
```

### Estimating contaminating fraction (automated)

Can do this manually by specifying a set of "non expressed genes" that should NOT be expressed in some cell types/clusters, or can do it automatically.

"The key thing to understand is that the contamination fraction estimate is the fraction of your data that will be discarded. If this value is set too low, your"corrected" data will potentially still be highly contaminated. If you set it too high, you will discard real data, although there are good reasons to want to do this at times (see section below). If the contamination fraction is in the right ball park, SoupX will remove most of the contamination. It will generally not matter if this number if off by a few percent."

For now, do automated estimation.

Contamination fraction set to 15% to remove \>98% of the contamination.

```{r automatedEstimation}
# set contamination fraction
list_soup = lapply(list_soup, setContaminationFraction, contFrac=0.15)

# automated contamination estimation
list_soup = lapply(list_soup, autoEstCont)
```

### Correcting expression profile

Method = subtraction

```{r correction}
out = lapply(list_soup, adjustCounts, roundToInt = TRUE)
#qsave(out,"rdata/00_out.qs", nthreads = 16)

# create seurat
seur_unfilt = map2(out, names(out), ~ CreateSeuratObject(.x, project = .y))
```

### QC metrics

-   UMI count (nCount_RNA)
-   Gene features count (nFeature_RNA)
-   Number of genes detected per UMI for complexity, see novelty score (log10GenesperUMI)
-   Perc. of mitochondrial reads (ratio_mt)
-   to do: calc number of reads w/ TSO, need to look into fastq files

```{r add to metadata}
# calc ratio of UMI/gene
calc_noveltyScore = function(seuObj){
  seuObj$log10Genes_per_UMI = log10(seuObj$nFeature_RNA) / log10(seuObj$nCount_RNA)
  return(seuObj)
}

# calc ratio of MT genes
calc_mt = function(seuObj){
  seuObj$ratio_mt = PercentageFeatureSet(seuObj, pattern = "^MT-")/100
  return(seuObj)
} 

seur_unfilt = lapply(seur_unfilt, calc_noveltyScore)
seur_unfilt = lapply(seur_unfilt, calc_mt)
```

Basic visualization with box plots.

```{r plot all test2, include=T, warning=F}
# get list of metadata
## TO DO: troubleshoot this??
metadata = map2(
  seur_unfilt, 
  names(seur_unfilt),
  ~.x@meta.data %>%
    rownames_to_column(var = "cell_ID") %>%
    mutate(LN_ID = .y)
) 

metadata = do.call(rbind,metadata)

# also pivot for plotting purposes, possibly unneccessary!
metadata_piv = pivot_longer(
  metadata, 
  cols = c("nCount_RNA","nFeature_RNA","log10Genes_per_UMI","ratio_mt"),
  names_to = "metric",
  values_to = "value"
)

# boxplot, all
ggplot(metadata_piv, aes(LN_ID,value)) +
  geom_boxplot(aes(fill = LN_ID),
               alpha = 0.5,
               outlier.shape = NA) +
  theme(aspect.ratio = 4/3,
        axis.text.x = element_blank(),
        axis.ticks.x = element_blank()) +
  scale_y_log10(n.breaks = 8) +
  facet_grid(rows = vars(metric), scales = "free") + 
  labs(x = NULL, y = "Value")

ggsave("figures/01_01_gexQCunflt.png")
```

Individual density plots + cut offs TO DO: PLOT with IQR cutoffs!

```{r plots w cutoffs, include=T}
# plotting function
plot_ridges = function(metadata.df, metric){
  p_out = ggplot(metadata.df,aes(.data[[metric]],fct_rev(LN_ID))) +
    geom_density_ridges(aes(fill = LN_ID), alpha = 0.5) +
    theme(aspect.ratio = 3/3,
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          panel.grid = element_blank(),
          axis.title.y = element_blank(),
          plot.title = element_text(hjust = 0.5),
          plot.subtitle = element_text(hjust = 0.5)) +
    scale_x_log10() +
    labs(title = metric)
  
  return(p_out)
}

# nCount_RNA (UMIs)
p1 = plot_ridges(metadata,"nCount_RNA")+
  geom_vline(xintercept = c(500,40000), alpha = 0.5) +
  labs(subtitle = "[500,40000]")

# nFeature_RNA (Unique Genes)
p2 = plot_ridges(metadata,"nFeature_RNA")+
  geom_vline(xintercept = c(300,7000), alpha = 0.5) +
  labs(subtitle = "[300,7000]")

# complexity
p3 = plot_ridges(metadata,"log10Genes_per_UMI")+
  geom_vline(xintercept = 0.85, alpha = 0.5) +
  labs(subtitle = "[0.85,1.0]")

# mt reads
p4 = plot_ridges(metadata,"ratio_mt")+
  geom_vline(xintercept = 0.15, alpha = 0.5) +
  scale_x_continuous() +
  xlim(c(0,0.3)) +
  labs(subtitle = "[0,0.15]")

# TO DO: remove more white space, save plot
ggpubr::ggarrange(
  p1,p2,p3,p4,
  nrow=2,ncol=2,
  common.legend = T, 
  legend = "right"
)

ggsave("figures/01_01_gexQC_density.png")
```

### Determining outliers by IQR

Use `nCount_RNA and nFeature_RNA < Q1-1.5IQR or > Q3 + 1.5IQR` to mark outlier cells.

```{r IQR calc, include=T}
get_q1 = function(x) quantile(x)[["25%"]]
get_q3 = function(x) quantile(x)[["75%"]]

# function to get which cells are outliers
find_outliers = function(seurObj){
  # get nCount_RNA and nFeatures_RNA
  nUMI = seurObj$nCount_RNA
  nGene = seurObj$nFeature_RNA
  
  # get names of cells which are outliers 
  nUMI_out = which(
    nUMI < (get_q1(nUMI) - 1.5*IQR(nUMI)) | nUMI > (get_q3(nUMI) + 1.5 * IQR(nUMI))
  )
  
  nGene_out = which(
    nGene < (get_q1(nGene) - 1.5*IQR(nGene)) | nGene > (get_q3(nGene) + 1.5 * IQR(nGene))
  )
  
  discardCells = union(names(nUMI_out),names(nGene_out))
  return(discardCells)
}

# get names of cells to DISCARD
toDISCARD = lapply(seur_unfilt,find_outliers)
```

### Filtering cells on MERGED dataset

Used the following filtering parameters:

\* cells are not outliers

\* cells meet minimum threshhold for UMI and Feature count:

\* 500 \<= nCount_RNA

\* 300 \<= nFeature_RNA

\* complexity score \>= 0.85

\* ratio_mt \<= 0.15

```{r initial filtering, include=T}
# create cell ids column
# seur_unfilt = lapply(seur_unfilt, function(x){
#   x$cell_ids = colnames(x)
#   return(x)
# })

# need function for map2 to work within subset function
'%ni%' <- function(x,y)!('%in%'(x,y))

# subset cells
seur_filt = map2(
  seur_unfilt, 
  toDISCARD, 
  ~ subset(
    .x,
    subset = (nCount_RNA >= 500) &
      (nCount_RNA <= 50000) &
      (Cells(.x) %ni% .y) &
      (log10Genes_per_UMI >= 0.85) &
      (ratio_mt <= 0.15)
  )
)
```

### Remove doublets with scDblFinder

-   Needed to remove low quality cells first (ie: low read counts / features)
-   Assumes rate of doublets predicted by 10X ("By default this is assumed to be 1% per thousand cells captured (so 4% among 4000 thousand cells), which is appropriate for 10x datasets. Corrections for homeotypic doublets will be performed on the given rate.")

```{r scDblFinder}
# convert to SingleCellExperiment class
seur_sce = lapply(seur_filt, as.SingleCellExperiment)

# find doublets without clustering
seur_sce = map(seur_sce, ~ scDblFinder(.x, clusters = F))

# filter cells based on predicted singlet or doublet
# check that all(colnames(seur_sce[[1]]) == colnames(seur_filt[[1]]))
seur_filt = lapply(seq_along(seur_filt), function(x){
  filt2 = seur_filt[[x]][,seur_sce[[x]]$scDblFinder.class == "singlet"]
  return(filt2)
})

names(seur_filt) = ln_id
```

Number of cells before and after initial filtering:

```{r nCells, include=T}
# create tibble for plotting, ncol = cells, nrow = features
df1 = tibble(
  count = sapply(seur_unfilt,ncol),
  filt = "Before",
  LN_ID = ln_id
)

df2 = tibble(
  count = sapply(seur_filt,ncol),
  filt = "After",
  LN_ID = ln_id
)

# plot
# TO DO: add text above
bind_rows(df1,df2) %>%
  mutate(filt = factor(filt, levels = c("Before","After"))) %>%
  ggplot(., aes(LN_ID, count, fill = filt)) + 
  geom_bar(stat = "summary",
           position="dodge") +
  theme(panel.grid = element_blank(),
        aspect.ratio = 3/3,
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Filtering based on RNA library", 
       y = "nCells", fill = "Filtered", x = NULL)

ggsave("figures/01_01_rnaFilt_summary.png",width=5,height = 3)
```

Before downstream steps (ie: normalization, scaling, annotation, actual analysis, etc), need to QC cells based on ATAC seq by using ArchR.
Afterwards, cells/barcodes that are still present after filtering each modality individual will be used for next steps.

```{r save rds, include=F}
qs::qsave(seur_filt, "rdata/01_seur_filt.qs", nthreads = 16)
```
